Rates of true polar wander in convecting planets

### 2015-08-27 tpw_rate.tex
comments

\begin{equation}
\frac{d {\bf H}}{dt} + {\bf \Omega} \times {\bf H} = 0
\label{eq:euler}
\end{equation}
known as Euler's equations, where ${\bf H} = {\bf I} \cdot {\bf \Omega}$ is the angular momentum vector, $\bf I$ is the moment of inertia tensor, and $\bf \Omega$ is the angular velocity.
On a dynamic planet $\bf I$ may be a function of time, so to conserve angular momentum $\bf \Omega$ must be also vary with time.
In this case Equation \eqref{eq:euler} is known as the Liouville equation \citep[e.g.][]{munk1960rotation}.
For a slowly convecting fluid such as a planetary mantle the inertial term $\partial {\bf H} / \partial t$ is negligible, so we may solve the simplified quasistatic equations
\begin{equation}
{\bf \Omega } (t)\times ( {\bf I}(t) \cdot {\bf \Omega }(t) ) = 0
\label{eq:quasistatic_liouville}
\end{equation}
Note that solving Equation \eqref{eq:quasistatic_liouville} is equivalent to solving an eigenvalue problem for ${\bf I}$, where the eigenvectors correspond to the principal axes and the eigenvalues correspond to the principal moments (where the most stable orientation of the planet corresponds to rotating about the largest principal axis).
In practice this eigenvalue approach has been often used in previous studies for computing the spin axis of a planet \citep[e.g.][]{steinberger1997changes, roberts2007cause}.
The moment of inertia tensor in \eqref{eq:quasistatic_liouville} includes all contributions to the mass structure
of the planet, including the spherically symmetric mass distribution, rotational deformation, deformation due to self gravity, internal and surface density anomalies, and surface deflections due to density anomalies.
Here we are interested in the moment of inertia due to mantle convection, and thus neglect the contributions of surface
density anomalies (though these can be quite important, \hl{ citations for Glaciers, Tharsis, etc} ).

@@

\begin{equation}
\begin{aligned}
- \nabla P_0 =  \rho_0 {\bf g} -  \rho_0 {\bf \Omega \times \Omega \times r}
\label{eq:hydrostatic}
\end{aligned}
\end{equation}

Nondimensionalizing with the parameters in Table \ref{tab:nondim} and removing the reference state we find the nondimensional momentum equation:
\begin{equation}
\begin{aligned}
 - \nabla P^* + \; \nabla^2{\bf v} - \mathrm{Ra} \; T \; {\bf g} + \mathrm{Ra \; Fr}\; T \;{\mitbf{\omega} \times \mitbf{\omega} \times {\bf r}} = 0
\end{aligned}
\end{equation}

Equation \eqref{eq:navier_stokes} can be transformed into an angular momentum equation by crossing with $\bf r$ and integrating over the volume of the mantle:

\begin{equation}
\begin{aligned}
-\int_V {\bf r} \times \nabla P \,dV + \int_V \eta {\bf r} \times \nabla^2 {\bf u} \,dV - \int_V \rho {\bf r \times g} \,dV 
   + \int_V \rho {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV &= 0 \\
\end{aligned}
\end{equation}

@@

\begin{equation}
\begin{aligned}
\int_V \rho {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV &= 0 \\
\end{aligned}
\end{equation}

which is a statement that a quasistatic body will rotate around the principle axis of its total moment of inertia.

Splitting the volume into $V_H$ and $\Delta V$ and substituting \eqref{eq:eos} we find

\begin{equation}
\begin{aligned}
&\int_{V_H} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV + 
\int_{V_H} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV +  \\
&\int_{\Delta V} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV + 
\int_{\Delta V} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV = 0  \\
\end{aligned}
\end{equation}

The first term of this equation is zero due to the hydrostatic equation.  
The fourth term is negligible due to being second order in the smallness parameters $\Delta V/V_H$ and $\Gamma$.
Removing these, we find
\begin{equation}
\begin{aligned}
\int_{\Delta V} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV = 
-\int_{V_H} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV
\end{aligned}
\end{equation}

This equation may be identified with \eqref{eq:disequilibrium}, where disequilibrium in the rotational deformation (left side) is balanced by the mismatch of the convective moment of intertia with the spin axis (right side).
Our goal is to identify characteristic sizes of these quantities, which must be functions of the nondimensional numbers identified in Table \ref{tab:nondim}.

@@

\begin{equation}
T_{\mathrm{degree 2}}(t) \sim \mathrm{Ra}^{-1}
\label{eq:degree_two_of_ra}
\end{equation}
For two-dimensional simulations shown below there is a reduced dimensionality when calculating the number of modes,
so $N_\mathrm{modes} \sim \left(R/d \right)^2$.  This leads us to a scaling of $T_{\mathrm{degree2}} \sim \mathrm{Ra}^{-2/3}$, which is used in the simulations shown in Figure~\ref{fig:eigengap}.
This result has a simple interpretation.
As the Rayleigh number of the system increases, the smallest lengthscale of convective features gets smaller.
The total power in the temperature field is spread across a larger spectrum, leaving less total power for the degree-two part, which is what drives TPW.

@@

The preceding results clarify the complex relationship between the Rayleigh number, the Froude number and the rates of TPW for a convecting planet.
As mantle convection redistributes mass in the planet's interior, the rotational bulge moves around to stay aligned with the principle axes of the convective moment of inertia. 
There is a constant competition between growth of the mismatch angle $\theta$ through Equation \eqref{eq:grow_perturbation} and its relaxation through Equation \eqref{eq:simple_milankovitch}.
We may plug in the estimate for $\Lambda_{ij}$ into equation \eqref{eq:simple_milankovitch} to find the strikingly simple expression
\begin{equation}
\frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{1}{\mathrm{Fr}} \sin{2 \theta}
\label{eq:simplest_milankovitch}
\end{equation}
Surprisingly, $\Gamma$ and $\mathrm{Ra}$ have completely dropped from the prefactor in the scaling. 
At high Rayleigh numbers response timescales for relaxation of the mismatch angle go down.
At the same time, however, coherence in the temperature structure goes down, reducing the amount of power in the degree-two part of the field responsible for driving TPW.
That these two effects cancel is something of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
Scalings for lengthscales of convection in fluids with temperature dependent viscosity \citep[e.g.][]{solomatov1995scaling} or pseudoplastic rheology \citep[e.g.][]{korenaga2010scaling} have different functional dependencies on $\mathrm{Ra}$ or additional nondimensional parameters.
However, a common feature in most scalings is that typical lengthscales are still some power-law of Rayleigh number $d \propto \sim \mathrm{Ra}^{-\beta}$.
With this form, our scaling for TPW rate has the following dependence on $\mathrm{Ra}$:
\begin{equation}
\dot{\theta} \propto \mathrm{Ra}^{1-3\beta}
\end{equation}
In general, $\beta$ is some small number between one-fourth and one-third, so we expect that more complicated estimates for $d(\mathrm{Ra})$ will still show a weak dependence on the Rayleigh number.

We then suggest that the most important parameters are the Froude number, which acts as the brakes on the system, and $\theta(\mathrm{Ra})$.
Whereas the rest of the expression has at most a weak dependence on $\mathrm{Ra}$, this is expected to be strongly dependent on it.

>>>>

\begin{equation}
\frac{d {\bf H}}{dt} + {\bf \Omega} \times {\bf H} = 0
\label{eq:euler}
\end{equation}
known as Euler's equations, where ${\bf H} = {\bf I} \cdot {\bf \Omega}$ is the angular momentum vector, $\bf I$ is the moment of inertia tensor, and $\bf \Omega$ is the angular velocity.
On a dynamic planet $\bf I$ may be a function of time, so to conserve angular momentum $\bf \Omega$ must be also vary with time.
In this case Equation \eqref{eq:euler} is known as the Liouville equation \citep[e.g.][]{munk1960rotation}.
For a slowly convecting fluid such as a planetary mantle the inertial term $\partial {\bf H} / \partial t$ is negligible, so we may solve the simplified quasistatic equations
\begin{equation}
{\bf \Omega } (t)\times ( {\bf I}(t) \cdot {\bf \Omega }(t) ) = 0
\label{eq:quasistatic_liouville}
\end{equation}
Equation \eqref{eq:quasistatic_liouville} indicates that ${\bf \Omega}$ and ${\bf I}$ are parallel, so a solution for ${\bf \Omega}(t)$ is equivalent to solving an eigenvalue problem for ${\bf I}(t)$, where the eigenvectors correspond to the principal axes and the eigenvalues correspond to the principal moments (where the most stable orientation of the planet corresponds to rotating about the largest principal axis).
In practice this eigenvalue approach has been often used in previous studies for computing the spin axis of a planet \citep[e.g.][]{steinberger1997changes, roberts2007cause}.
The moment of inertia tensor in \eqref{eq:quasistatic_liouville} includes all contributions to the mass structure
of the planet, including the spherically symmetric mass distribution, rotational deformation, deformation due to self gravity, internal and surface density anomalies, and surface deflections due to density anomalies.
Here we are interested in contributions from mantle convection, so we neglect other processes.

@@

\begin{equation}
\begin{aligned}
- \nabla P_0 =  \rho_0 {\bf g} -  \rho_0 {\bf \Omega \times \Omega \times r}
\label{eq:hydrostatic}
\end{aligned}
\end{equation}
Nondimensionalizing with the parameters in Table \ref{tab:nondim} and removing the reference state we find the nondimensional momentum equation:
\begin{equation}
\begin{aligned}
 - \nabla P^* + \; \nabla^2{\bf v} - \mathrm{Ra} \; T \; {\bf g} + \mathrm{Ra \; Fr}\; T \;{\mitbf{\omega} \times \mitbf{\omega} \times {\bf r}} = 0
\end{aligned}
\end{equation}
Equation \eqref{eq:navier_stokes} can be transformed into an angular momentum equation by crossing with $\bf r$ and integrating over the volume of the mantle:
\begin{equation}
\begin{aligned}
-\int_V {\bf r} \times \nabla P \,dV + \int_V \eta {\bf r} \times \nabla^2 {\bf u} \,dV - \int_V \rho {\bf r \times g} \,dV 
   + \int_V \rho {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV &= 0 \\
\end{aligned}
\end{equation}

@@

\begin{equation}
\begin{aligned}
\int_V \rho {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV &= 0 \\
\end{aligned}
\end{equation}
which is a statement that a quasistatic body will rotate around the principle axis of its total moment of inertia.

Splitting the volume into $V_H$ and $\Delta V$ and substituting \eqref{eq:eos} we find
\begin{equation}
\begin{aligned}
&\int_{V_H} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV + 
\int_{V_H} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV +  \\
&\int_{\Delta V} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV + 
\int_{\Delta V} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV = 0  \\
\end{aligned}
\end{equation}
The first term of this equation is zero due to the hydrostatic equation.  
The fourth term is negligible due to being second order in the smallness parameters $\Delta V/V_H$ and $\Gamma$.
Removing these, we find
\begin{equation}
\begin{aligned}
\int_{\Delta V} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV = 
-\int_{V_H} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV
\end{aligned}
\end{equation}
This equation may be identified with \eqref{eq:disequilibrium}, where disequilibrium in the rotational deformation (left side) is balanced by the mismatch of the convective moment of intertia with the spin axis (right side).
Our goal is to identify characteristic sizes of these quantities, which must be functions of the nondimensional numbers identified in Table \ref{tab:nondim}.

@@

\begin{equation}
T_{\mathrm{degree 2}}(t) \sim \mathrm{Ra}^{-1}
\label{eq:degree_two_of_ra}
\end{equation}

For two-dimensional simulations shown below there is a reduced dimensionality when calculating the number of modes,
so $N_\mathrm{modes} \sim \left(R/d \right)^2$.  This leads us to a scaling of $T_{\mathrm{degree2}} \sim \mathrm{Ra}^{-2/3}$, which is used in the simulations shown in Figure~\ref{fig:eigengap}.
This result has a simple interpretation.
As the Rayleigh number of the system increases, the smallest lengthscale of convective features gets smaller.
The total power in the temperature field is spread across a larger spectrum, leaving less total power for the degree-two part, which is what drives TPW.

@@

The preceding results clarify the complex relationship between the Rayleigh number, the Froude number and the rates of TPW for a convecting planet.
As mantle convection redistributes mass in the planet's interior, the rotational bulge moves around to stay aligned with the principle axes of the convective moment of inertia. 
There is a constant competition between growth of the mismatch angle $\theta$ through Equation \eqref{eq:grow_perturbation} and its relaxation through Equation \eqref{eq:simple_milankovitch}.
We may plug in the estimate for $\Lambda_{ij}$ into equation \eqref{eq:simple_milankovitch} to find the strikingly simple expression
\begin{equation}
\frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{1}{\mathrm{Fr}} \sin{2 \theta}
\label{eq:simplest_milankovitch}
\end{equation}
Surprisingly, $\Gamma$ and $\mathrm{Ra}$ have completely dropped from the prefactor in the scaling. 
At high Rayleigh numbers response timescales for relaxation of the mismatch angle go down.
At the same time, however, coherence in the temperature structure goes down, reducing the amount of power in the degree-two part of the field responsible for driving TPW.
That these two effects cancel is something of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
Scalings for lengthscales of convection in fluids with temperature dependent viscosity \citep[e.g.][]{solomatov1995scaling} or pseudoplastic rheology \citep[e.g.][]{korenaga2010scaling} have different functional dependencies on $\mathrm{Ra}$ or additional nondimensional parameters.
However, a common feature in most scalings is that typical lengthscales are still some power-law of Rayleigh number $d \propto \sim \mathrm{Ra}^{-\beta}$.
With this form, our scaling for TPW rate has the following dependence on $\mathrm{Ra}$:
\begin{equation}
\frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{\mathrm{Ra}^{1-3\beta}}{\mathrm{Fr}} \sin{2 \theta}
\end{equation}
In general, $\beta$ is some small number between one-fourth and one-third, so we expect that more complicated estimates for $d(\mathrm{Ra})$ will still show a weak dependence on the Rayleigh number.

We then suggest that the most important parameters are the Froude number, which acts as the brakes on the system, and the mismatch angle $\theta(\mathrm{Ra})$.
Whereas the rest of the expression has at most a weak dependence on $\mathrm{Ra}$, $\theta$ is expected to be strongly dependent on it.
